Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2021-12-23 manuscripts/owp_fim4_2021/sections/discussion.tex
minor edits

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's N values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V[y] or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B[y] or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively. 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but since both are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ is plotted against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

Our qualitative analysis suggests that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the NED 10m DEM is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's N which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3m and 1m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

>>>>

In analyzing the data, we found a slight reduction in FAR was detected and more digging pointed to a bias in rating curves introduced by stream order reduction.
Figure \ref{fig:rating_curve_comparison} illustrates the general effect that stream order reduction has on synthetic rating curves.
Sub-figure \ref{fig:rating_curve_comparison}a shows how the average rating curves for all reaches for stage values 0 to 25 meters at one-third meter intervals tend to bias down (and to the right) with ever increasing stream order reduction (FR to MS to GMS). 
This bias is more pronounced for GMS since that implements stream order reduction down to the unit level for the entire FR network while MS only does so for 4-5\% of the network.
Attempting to diagnose this bias in the SRC leads one to Equation \ref{eq:reach_averaged_mannings_equation} which shows the reach averaged synthetic rating curve relationship between stage and discharge.
Across the three methods explored, FR, MS, and GMS, one identifies differences in the inputs and outputs and notes no difference in the stages and Manning's n values.
While the channel slope and reach lengths are not exactly the same across methods, their averaged differences are very negligible which only leaves room for deviations in volume and bed area.
Again, volume (V[y] or simply V) is synonymous to reach-averaged cross-sectional area and bed area (B[y] or B) is analogous to reach-averaged hydraulic radius.
Discharge, Q, is directly related to volume and inversely related to bed area and each parameter is weighed according to the magnitude of its exponent which are $\frac{5}{3}$ and $\frac{2}{3}$ respectively. 
Figures \ref{fig:rating_curve_comparison} b and c show how volume and bed area compare across the three methods with GMS having significantly greater values than MS which has greater values than FR.
Again the relative discrepancy between FR vs MS and MS vs GMS is explained by the extent of their spatial coverages.
Both V and B values increase but since both are weighed differently by their respective exponents and pull Q in different directions.
We show in Figure \ref{fig:rating_curve_comparison}d the relationship of $\frac{V^{5/3}}{B^{2/3}}$ is plotted against stage, y, to show how these two parameters collectively pull the rating curve Q up and biases the rating curve down.
In other words, the magnitude and weight of the volume at each stage level exceeds the influence of the magnitude and weight of the bed area.
Both parameters are set to increase mainly due to much larger catchments leading to more pixels at each stage level as shown in Figure \ref{fig:rating_curve_comparison}e.
Much of the increase in inundated pixels, volume, and bed area can be explained by much larger catchments that encompass neighboring tributaries.
These tributaries have a significant amount of bathymetry that is low-lying thus easily including the SRC derivation. 
They also contribute volume and bed area that is technically not perpendicular to the flux of streamflow being accounted for in the stream in question. 
Careful examination of Figure \ref{fig:gms_enhancement}b shows how much larger catchments include neighboring tributaries and the geometry associated with those tributaries. 
This geometry is not perpendicular to the flow that is associated with the main reach thus leading to biases in the SRC.
We consider this fact to have a nuanced effect on skill, while reducing the rate of FPs it also can lead to FNs due to biases in the SRC.

@@

Our qualitative analysis suggests that the synthetic rating curves offer a significant opportunity for improvement in HAND based FIM for future development.
The bathymetry of the NED 10m DEM is known to be lacking proper representation thus leading to inadequate representation of volume and bed area with all three methods employed.
Manning's n which typically accounts for roughness could be tuned to account for these DEM limitations or could be held fixed to some local value associated with a given flood magnitude.
Some adjusting parameter must be introduced to enhance the estimation of the bathymetric representation.
Lidar DEMs from the USGS at 3m and 1m scale could be utilized to derive HAND as well which we conject should show better agreement with higher fidelity FIMs also derived from the same Lidar based DEMs.

### 2021-12-23 manuscripts/owp_fim4_2021/sections/materials_and_methods.tex
minor edits

where Q is discharge, y indicates the stage, n is the Manning's N roughness coefficient, V is volume at stage y, S is channel slope, L is along flow length, and B is wetted bed area at stage y.

@@

Setting of the Manning's N roughness coefficient has precedent in previous CFIM studies \cite{maidment2017conceptual,liu2016cybergis,liu2020height,djokic2019arc,garousi2019terrain,zheng2018geoflood} with two noted values of 0.05 and 0.06 for NFIE and \citeA{djokic2019arc} respectively. 

@@

We note significant opportunity to enhance CFIM skill by better parameterizing Manning's N according to available data including but limited to land cover, land use, stream order, stream geometry, drainage area, reach length, and discharge percentiles \cite{garousi2019terrain,johnson2019integrated}.
For now and for the purpose of this study, we examine the developed ecosystem of tools with Manning's N set to both 0.06 and 0.12 which we hope will shed some light on the sensitivity of this parameter to HAND based FIMs.

>>>>

where Q is discharge, y indicates the stage, n is the Manning's n roughness coefficient, V is volume at stage y, S is channel slope, L is along flow length, and B is wetted bed area at stage y.

@@

Setting of the Manning's n roughness coefficient has precedent in previous CFIM studies \cite{maidment2017conceptual,liu2016cybergis,liu2020height,djokic2019arc,garousi2019terrain,zheng2018geoflood} with two noted values of 0.05 and 0.06 for NFIE and \citeA{djokic2019arc} respectively. 

@@

We note significant opportunity to enhance CFIM skill by better parameterizing Manning's n according to available data including but limited to land cover, land use, stream order, stream geometry, drainage area, reach length, and discharge percentiles \cite{garousi2019terrain,johnson2019integrated}.
For now and for the purpose of this study, we examine the developed ecosystem of tools with Manning's n set to both 0.06 and 0.12 which we hope will shed some light on the sensitivity of this parameter to HAND based FIMs.

### 2021-12-23 manuscripts/owp_fim4_2021/sections/results.tex
minor edits

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We produced FIMs for the entire BLE domain within the 49 HUC8 area across several states in the south central US. 
The forecasted FIMs using the discharges for the 1\% (100 year) and 0.2\% (500 year) recurrence flows directly from HEC-RAS were used to avoid noise and errors from hydrological processes.
We computed the statistics (CSI, POD, and FAR) for both 100 and 500 year events for Mannings N set to 0.06 and 0.12. 
The distribution of these statistics can be examined in Figure \ref{fig:violin_plot} as violin plots.
Each half of a violin plot represents the kernel density estimation (KDE) for a given model (FR, MS, GMS), given Manning's N value (0.06, 0.12), and given recurrence interval (1\%, 0.2\%), and performance metric (CSI, POD, FAR).
We also denote trend lines for each metric and Manning's n setting as well as their respective slope estimate and one-tailed p-value denoting the level of significance of the trend.

Aggregating the metrics in the method above treats each HUC8 as it's own unit and does little to consider the size differences of the HUCs. 
In an opposing aggregation method, we illustrate in Table \ref{tab:aggregate_metrics} the sum of all the TPs, FPs, and FNs to recompute CSI, POD, and FAR across the entire domain. 
%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.9]{figures/violin_plots.jpg}
\caption{Shows kernel density estimation of the distributions (sample size = 49) for 1\% (100 year) and 0.2\% (500 year) along with horizontal, dashed lines for the 25th, 50th, and 75th percentiles (in order from bottom to top).
The sub-figures separate the combination of three metrics (CSI, POD, and FAR) for two settings of Manning's N (0.06 and 0.12).
Trend lines for each metric - Mannings combination are shown (sample size = 294) along with associated slope and p-value of slope testing one-tailed significance.}
\label{fig:violin_plot}
\end{figure}
%
\begin{table}[h!]
\caption{Primary metrics, TPs, FPs, and FNs, aggregated for BLE domain to recompute CSI, POD, and FAR.}
\label{tab:aggregate_metrics}
\centering
%\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
\begin{tabular}{|c|c||c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Metric} & \multirow{2}{*}{Manning's N} & \multicolumn{2}{|c|}{FR} & \multicolumn{2}{|c|}{MS} & \multicolumn{2}{|c|}{GMS} \\
\cline{3-8}
  &  & 100yr & 500yr & 100yr & 500yr & 100yr & 500yr \\
\hline
\multirow{2}{*}{CSI} & 0.06 & 0.5576 & 0.5839 & 0.5717 & 0.5990 & 0.5796 & 0.6075 \\
\cline{2-8}
  & 0.12 & 0.5915 & 0.6149 & 0.6054 & 0.6288 & 0.6182 & 0.6435 \\
\hline
\multirow{2}{*}{POD} & 0.06 & 0.6354 & 0.6575 & 0.6524 & 0.6755 & 0.6633 & 0.6863 \\
\cline{2-8}
  & 0.12 & 0.7255 & 0.7446 & 0.7460 & 0.7648 & 0.7606 & 0.7810 \\
\hline
\multirow{2}{*}{FAR} & 0.06 & 0.1800 & 0.1609 & 0.1778 & 0.1589 & 0.1787 & 0.1589 \\
\cline{2-8}
  & 0.12 & 0.2379 & 0.2208 & 0.2374 & 0.2204 & 0.2324 & 0.2148 \\
\hline
\end{tabular}
\end{table}
%
% Interpretation of metrics
From Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics}, we denote several meaningful trends. 
Using CSI as an overall proxy for skill of the FIM, we note that generally speaking the skill is correlated with a reduction of the stream orders of the processing units used for HAND.
In other words, the more you derive HAND on networks of unit drainage density and mosaic the resulting FIMs, the better those FIMs perform.
While this trend is evident for both sets of Manning's N values, the trend is slightly more significant for the higher value of 0.12.
Other trends related to this Figure include the general performance premium for 0.2\% events as opposed to lower 1\% events.
We also note how the higher Manning's N value enhances performance for both of these recurrence intervals across all models.

Dissecting the improvements and trends presented in the previous paragraph comes down mostly to improvement in POD or a reduction in absolute amount of FNs.
POD being the primary driver in skill enhancement is evident across models by comparing the slope of the POD lines with the slope of the FAR lines.
Even though aggregating metrics by HUC8 yields a statistically zero trend, one does see a slight reduction in FAR across models that reduce HAND's maximum stream order.
Additionally, we note that POD is a primary driver in enhancing performance across Manning's N values as well.
This significant improvement comes at a cost of false alarms as the FAR increases significantly across Manning's N values.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computational Performance}
\label{ssec:compuational_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The NFIE experiments were able to produce HAND for 331 HUC6's in 1.34 CPU years \cite{liu2016cybergis} and estimates using work from \citeA{djokic2019arc} put producing HAND at the FR NWM at 0.55 CPU years. 
For our work, we were able to produce HAND at the full NWM resolution in 0.13 CPU years which represents a substantial speed-up compared to previous works.
For the MS resolution an additional, 0.05 CPU years is required on top of this bringing the total to about 0.18 CPU years to produce 2,188 HUC8s that span additional areas not covered in previous HAND versions including Hawaii and Puerto Rico.
GMS which generalizes HAND production to level path scale adds a significant amount of CPU time to the process bringing the estimate total to about 1.17 CPU years.
%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We produced FIMs for the entire BLE domain within the 49 HUC8 area across several states in the south central US. 
The forecasted FIMs using the discharges for the 1\% (100 year) and 0.2\% (500 year) recurrence flows directly from HEC-RAS were used to avoid noise and errors from hydrological processes.
We computed the statistics (CSI, POD, and FAR) for both 100 and 500 year events for Mannings N set to 0.06 and 0.12. 
The distribution of these statistics can be examined in Figure \ref{fig:violin_plot} as violin plots.
Each half of a violin plot represents the kernel density estimation (KDE) for a given model (FR, MS, GMS), given Manning's n value (0.06, 0.12), and given recurrence interval (1\%, 0.2\%), and performance metric (CSI, POD, FAR).
We also denote trend lines for each metric and Manning's n setting as well as their respective slope estimate and one-tailed p-value denoting the level of significance of the trend.

Aggregating the metrics in the method above treats each HUC8 as it's own unit and does little to consider the size differences of the HUCs. 
In an opposing aggregation method, we illustrate in Table \ref{tab:aggregate_metrics} the sum of all the TPs, FPs, and FNs to recompute CSI, POD, and FAR across the entire domain. 
%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.9]{figures/violin_plots.jpg}
\caption{Shows kernel density estimation of the distributions (sample size = 49) for 1\% (100 year) and 0.2\% (500 year) along with horizontal, dashed lines for the 25th, 50th, and 75th percentiles (in order from bottom to top).
The sub-figures separate the combination of three metrics (CSI, POD, and FAR) for two settings of Manning's n (0.06 and 0.12).
Trend lines for each metric - Mannings combination are shown (sample size = 294) along with associated slope and p-value of slope testing one-tailed significance.}
\label{fig:violin_plot}
\end{figure}
%
\begin{table}[h!]
\caption{Primary metrics, TPs, FPs, and FNs, aggregated for BLE domain to recompute CSI, POD, and FAR.}
\label{tab:aggregate_metrics}
\centering
%\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
\begin{tabular}{|c|c||c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Metric} & \multirow{2}{*}{Manning's n} & \multicolumn{2}{|c|}{FR} & \multicolumn{2}{|c|}{MS} & \multicolumn{2}{|c|}{GMS} \\
\cline{3-8}
  &  & 100yr & 500yr & 100yr & 500yr & 100yr & 500yr \\
\hline
\multirow{2}{*}{CSI} & 0.06 & 0.5576 & 0.5839 & 0.5717 & 0.5990 & 0.5796 & 0.6075 \\
\cline{2-8}
  & 0.12 & 0.5915 & 0.6149 & 0.6054 & 0.6288 & 0.6182 & 0.6435 \\
\hline
\multirow{2}{*}{POD} & 0.06 & 0.6354 & 0.6575 & 0.6524 & 0.6755 & 0.6633 & 0.6863 \\
\cline{2-8}
  & 0.12 & 0.7255 & 0.7446 & 0.7460 & 0.7648 & 0.7606 & 0.7810 \\
\hline
\multirow{2}{*}{FAR} & 0.06 & 0.1800 & 0.1609 & 0.1778 & 0.1589 & 0.1787 & 0.1589 \\
\cline{2-8}
  & 0.12 & 0.2379 & 0.2208 & 0.2374 & 0.2204 & 0.2324 & 0.2148 \\
\hline
\end{tabular}
\end{table}
%
% Interpretation of metrics
From Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics}, we denote several meaningful trends. 
Using CSI as an overall proxy for skill of the FIM, we note that generally speaking the skill is correlated with a reduction of the stream orders of the processing units used for HAND.
In other words, the more you derive HAND on networks of unit drainage density and mosaic the resulting FIMs, the better those FIMs perform.
While this trend is evident for both sets of Manning's n values, the trend is slightly more significant for the higher value of 0.12.
Other trends related to this Figure include the general performance premium for 0.2\% events as opposed to lower 1\% events.
We also note how the higher Manning's n value enhances performance for both of these recurrence intervals across all models.

Dissecting the improvements and trends presented in the previous paragraph comes down mostly to improvement in POD or a reduction in absolute amount of FNs.
POD being the primary driver in skill enhancement is evident across models by comparing the slope of the POD lines with the slope of the FAR lines.
Even though aggregating metrics by HUC8 yields a statistically zero trend, one does see a slight reduction in FAR across models that reduce HAND's maximum stream order.
Additionally, we note that POD is a primary driver in enhancing performance across Manning's n values as well.
This significant improvement comes at a cost of false alarms as the FAR increases significantly across Manning's n values.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computational Performance}
\label{ssec:compuational_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The NFIE experiments were able to produce HAND for 331 HUC6's in 1.34 CPU years \cite{liu2016cybergis} and estimates using work from \citeA{djokic2019arc} put producing HAND at the FR NWM at 0.55 CPU years. 
For our work, we were able to produce HAND at the full NWM resolution in 0.13 CPU years which represents a substantial speed-up compared to previous works.
For the MS resolution an additional, 0.05 CPU years is required on top of this bringing the total to about 0.18 CPU years to produce 2,188 HUC8s that span additional areas not covered in previous HAND versions including Hawaii and Puerto Rico.
GMS which generalizes HAND production to level path scale adds a significant amount of CPU time to the process bringing the estimate total to about 1.17 CPU years.
%
